---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
set.seed(42)
```

# SNITCH: Semi-supervised Non-linear Identification and Trajectory Clustering for High-dimensional Data

<!-- badges: start -->
<!-- badges: end -->

## Cite
If you're using SNITCH in your work, please cite:

> Grolaux R., *et al.* **SNITCH**: Semi-supervised Non-linear Identification and Trajectory Clustering for High-dimensional data. (preprint/manuscript link forthcoming)

---

## Overview

**SNITCH** (**S**emi-supervised **N**on-linear **I**dentification and **T**rajectory **C**lustering for **H**igh-dimensional data) is an R package to analyze **ageing-related DNA methylation trajectories**. It provides a robust, end-to-end workflow to:

- Classify CpG sites into **linear**, **non-linear (NL)**, or **non-correlated** trajectories.
- Identify **DMPs**, **VMPs**, and **non-linear DMPs** driven by age.
- Perform **FPCA** on **smoothed non-linear** trajectories to capture complex ageing patterns.
- **Cluster** non-linear CpGs with **k-means**, **MFUZZ (fuzzy)**, or **HDBSCAN**, and compare results via **ARI/AMI**.

SNITCH aims to be **efficient**, **scalable**, and **flexible**: Try it on your dataset!


---

## Installation

You can install the development version of SNITCH from GitHub with:

```r
# install.packages("pak")
pak::pak("fishrscale/SNITCH")
```

Or using **devtools**:

```r
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("fishrscale/SNITCH")
```

> The clustering demo below also uses: `dbscan`, `factoextra`, `ggrepel`, `aricode`, `dplyr`, `tibble`, `ggplot2`, and Bioconductor packages `Biobase`, `Mfuzz`.
> Install Bioconductor deps with:
> ```r
> if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
> BiocManager::install(c("Biobase", "Mfuzz"))
> ```

---

## Quick Start

### 1️⃣ Simulate DNA Methylation Data

Generate a dataset with simulated methylation values for multiple CpG sites following different ageing-related trajectories.

```{r quickstart-sim, eval=FALSE}
library(SNITCH)

# Simulate data with 300 individuals
simulated_data <- simulate_methylation_data(n_people = 300, plot = TRUE)

# Extract data frames
ages_df        <- simulated_data$ages
groups_df      <- simulated_data$groups
methylation_df <- simulated_data$meth
```

This will generate and save in a new "Results" folder within the current working directory:
- `functions_sim_data.pdf`: Plots of the predefined methylation patterns.
- `sample_sim_data.pdf`: Visualization of the simulated CpG sites.

### 2️⃣ Classify CpG Sites

Determine which CpG sites follow linear, non-linear, or non-correlated trajectories.

```{r quickstart-classify, eval=FALSE}
scaled_data <- prepare_data(data = t(methylation_df), age = ages_df$Age)

classified_cpgs <- run_parallel_classification(dat_scaled = scaled_data$dat_scaled, 
                                               age = scaled_data$Age, 
                                               ages_grid = scaled_data$ages_grid)

head(classified_cpgs)
```

### 3️⃣ Perform FPCA on Smoothed Non-Linear CpGs

```{r quickstart-fpca, eval=FALSE}
# Select only non-linear CpGs for FPCA
non_linear_cpgs <- classified_cpgs[grep("NL", classified_cpgs$classification),]$CpG
nl_smooth_rows  <- classified_cpgs[grep("NL", classified_cpgs$classification),]
nl_smooth       <- do.call(rbind, nl_smooth_rows$Predictions)
rownames(nl_smooth) <- non_linear_cpgs

# Perform FPCA
fpca_results <- perform_fpca(nl_var_smooth = nl_smooth, ages_grid = scaled_data$ages_grid)

# Plot FPCA results (files are saved by the function)
plot_fpca_results(fpca_results, ages_grid = scaled_data$ages_grid)
```

### 4️⃣ Cluster NL CpGs (k-means, MFUZZ, HDBSCAN) + Compare (ARI/AMI)

Use your favorite **unsupervised clustering** strategy to group non-linear CpGs using their **FPCA scores**. Below we illustrate three common options: **k-means**, **fuzzy clustering (MFUZZ)**, and **HDBSCAN**, and compare them with **Adjusted Rand Index (ARI)** and **Adjusted Mutual Information (AMI)** against the ground truth labels. Use the diagnostics to pick sensible parameters for your data.

```{r quickstart-cluster, eval=FALSE}
library(dbscan)     # HDBSCAN
library(Mfuzz)      # Fuzzy c-means
library(Biobase)    # ExpressionSet
library(factoextra) # k-means diagnostics
library(ggplot2)
library(ggrepel)
library(aricode)    # ARI/AMI
library(dplyr)
library(tibble)

# Ground truth from simulation
truth <- groups_df$Group

# We'll update copies of 'classified_cpgs' so we can compare methods cleanly
class_work <- classified_cpgs
nl_mask    <- class_work$classification == "NL"

# Container for method comparison
df_comp <- tibble(Method = character(), ARI = double(), AMI = double())

add_metrics <- function(pred_labels, method_label) {
  tibble(
    Method = method_label,
    ARI    = aricode::ARI(pred_labels, truth),
    AMI    = aricode::AMI(pred_labels, truth)
  )
}

# A) MFUZZ
eset <- new("ExpressionSet", exprs = fpca_results$scores)
m_opt <- mestimate(eset)
tmp   <- Dmin(eset, m_opt, crange = seq(2, 20, 1), repeats = 3, visu = TRUE)
ggplot2::ggsave("SNITCH_fuzzy_distance.pdf", width = 6, height = 4)

c_opt <- 11
mfuzz_fit <- mfuzz(eset, c = c_opt, m = m_opt)
fuzzy_assign <- apply(mfuzz_fit$membership, 1, which.max)
pred_fuzzy   <- paste0("NL_", fuzzy_assign)
cw_fuzzy <- class_work; cw_fuzzy$classification[nl_mask] <- pred_fuzzy
df_comp  <- dplyr::bind_rows(df_comp, add_metrics(cw_fuzzy$classification, "SNITCH + Fuzzy"))

# B) k-means
p_elbow <- fviz_nbclust(fpca_results$scores, kmeans, k.max = 20, method = "wss") +
  labs(title = "Elbow Method for K-Means",
       x = "Number of Clusters (K)",
       y = "Total Within-Cluster Sum of Squares (WCSS)") +
  theme_minimal()
ggplot2::ggsave("SNITCH_kmeans_elbow.pdf", p_elbow, width = 6, height = 4)

k_opt  <- 8
km_fit <- kmeans(as.matrix(fpca_results$scores), centers = k_opt, nstart = 25)
pred_km <- paste0("NL_", km_fit$cluster)
cw_km <- class_work; cw_km$classification[nl_mask] <- pred_km
df_comp <- dplyr::bind_rows(df_comp, add_metrics(cw_km$classification, "SNITCH + K-Means"))

# C) HDBSCAN
hdb_fit <- hdbscan(as.matrix(fpca_results$scores), minPts = 5)
pred_hdb <- paste0("NL_", hdb_fit$cluster)
cw_hdb <- class_work; cw_hdb$classification[nl_mask] <- pred_hdb
df_comp <- dplyr::bind_rows(df_comp, add_metrics(cw_hdb$classification, "SNITCH + HDBSCAN"))

# D) Compare methods
p_comp <- ggplot(df_comp, aes(x = ARI, y = AMI, label = Method)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3) +
  coord_equal() +
  labs(title = "Clustering Agreement on Simulated Data",
       x = "Adjusted Rand Index (ARI)",
       y = "Adjusted Mutual Information (AMI)") +
  theme_minimal()

print(df_comp)
print(p_comp)
ggplot2::ggsave("SNITCH_clustering_ari_ami.pdf", p_comp, width = 6, height = 4)
```

> **Why `eval = FALSE` above?**  
> The Quick Start sections mirror the full workflow but are turned off during README knit to keep it fast and avoid heavy optional dependencies. See the **Demo figures** below for runnable, lightweight chunks that generate images you can commit.

---

## Demo figures

### A) Example simulated trajectories

```{r demo-trajectories, fig.width=8, fig.height=5, echo=FALSE}
# Explicitly attach SNITCH so the knit has access to exported functions
library(SNITCH)
library(ggplot2); library(dplyr); library(tidyr); library(tibble)

# Smaller, faster sim for README
sim <- simulate_methylation_data(n_people = 150, n_sites = 2, plot = FALSE, output_dir = tempdir())
ages_df        <- sim$ages
groups_df      <- sim$groups
methylation_df <- sim$meth

demo <- methylation_df %>%
  rownames_to_column("Site") %>%
  pivot_longer(-Site, names_to = "Person", values_to = "Intensity") %>%
  mutate(Person = as.integer(Person)) %>%
  dplyr::left_join(ages_df,  by = "Person") %>%
  dplyr::left_join(groups_df, by = "Site")

keep_groups <- unique(demo$Group)[1:min(6, length(unique(demo$Group)))]
demo <- dplyr::filter(demo, Group %in% keep_groups)

ggplot(demo, aes(Age, Intensity, color = Group)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(se = FALSE, method = "loess", span = 0.6) +
  scale_y_continuous(limits = c(0, 1)) +
  facet_wrap(~ Group, ncol = 3) +
  labs(title = "Example simulated methylation trajectories",
       x = "Age", y = "DNA methylation (β)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

---

## Contributing
Issues and PRs are welcome! Please see the issue tracker: <https://github.com/fishrscale/SNITCH/issues>.

## License
Apache License 2.0. See `LICENSE` for details.

---

🚀 **SNITCH** — Bringing Non-Linear insights to your analysis.
